---
title: "International Debt"
author: "Nicole Rodriguez"
date: "2024-03-19"
output:
  pdf_document:
    toc: true
  html_document:
    theme: cosmo
    highlight: monochrome
    toc: true
    code_folding: show
---

International debt data collected by [The World Bank](https://www.worldbank.org/en/programs/debt-statistics) was analyzed. Throughout this analysis, the amount of debt (in USD) owed by developing countries is viewed and compared. 

```{r}
# Load the required libraries
library(tidyr)
library(dplyr)
library(DBI)
library(RSQLite)

# Read file from The World Bank
data <- read.csv("international_debt.csv")

# Rename columns 
colnames(data) <- c("Country_Name",
  "Country_Code",
  "External_Debt",
  "Debt_Value",
  "Total_Percent_Employment",
  "Annual_Population_Growth"
)

# Connect to SQLite database
connection <- dbConnect(RSQLite::SQLite(), "imdb.db")

# Write data to the database
dbWriteTable(connection, "international_data", data, overwrite = TRUE)
```


Across `r paste(nrow(data))` countries, the top 10 countries with the highest external debt were extracted. 

```{r}
# View data structure
head(data)
summary(data)

# Top 10 countries with the highest debt
top_10_debt <- dbGetQuery(connection, "
    SELECT Country_Name, External_debt, Debt_Value
    FROM international_data
    ORDER BY External_Debt DESC
    LIMIT 10;
")
# Return results 
print(top_10_debt)

# Calculate average national debt
avg_debt <- dbGetQuery(connection, 
  "SELECT AVG(External_Debt)
  FROM international_data;
")
# Return results
print(avg_debt)

```

The country with the highest debt was China with a whopping total external debt of $464,366,000,000 for the year of 2022. This is significantly higher than all the other countries. The average external debt is approximately $28,280,221,492.

```{r}
# Load required packages 
library(ggplot2)

# Bar plot; top 10 countries by debt
ggplot(top_10_debt, 
       aes(x = Country_Name, y = External_Debt, fill = Debt_Value)) +
  # Bar plot
  geom_bar(stat = "identity") + 
  # Add labels
  labs(title = "Top 10 Counties by Debt", x = "Country", y = "Total External Debt") + 
  # Additional plot structures/features 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))

```

Now we will look at the employment rates accross the countries and see if there is a correlation between these two factors. 

```{r}
# Top 10 countries with the highest percent of employment
top_10_employed <- dbGetQuery(connection, "
    SELECT Country_Name, ROUND(Total_Percent_Employment, 2) AS  Total_Percent_Employment
    FROM international_data
    ORDER BY Total_Percent_Employment DESC
    LIMIT 10
")
# Return results 
print(top_10_employed)

# Average percent employment 
avg_employment <- dbGetQuery(connection, 
  "SELECT ROUND(AVG(Total_Percent_Employment), 2) AS Total_Percent_Employment
  FROM international_data;
")
# Return results
print(avg_employment)

# Chinas employment percentage 
china_employment <- dbGetQuery(connection, 
  "SELECT Country_Name, ROUND(Total_Percent_Employment, 2) AS Total_Percent_Employment
  FROM international_data
  WHERE Country_Name = 'China';
")
# Return Results
print(china_employment)
```
The top 10 countries with the highest employment percentages are: 

- Qatar
- Kuwait
- Bahrain
- Oman
- United Arab Emirates	
- Norway
- Belarus
- United States	
- Saudi Arabia	
- Macao SAR, China	


Given that China's external debt surpasses all countries, I was curious about their employment rate and how it compared to the average employment rate. China's employment rate was `r paste(china_employment)` which isn't far from the overall average of `r paste(avg_employment)`. This made me wonder if there is any correlation between the total employment rate of countries and the total external debt they have. 

```{r}
# Load required libraries
library(dplyr)

# Debt/Employment variable
empl_debt <- dbGetQuery(connection, "
    SELECT External_Debt, Total_Percent_Employment
    FROM international_data;
")
# Remove NA values
empl_debt <- na.omit(empl_debt)

# Correlation test
empl_debt_corr <- cor(empl_debt$External_Debt, empl_debt$Total_Percent_Employment)

# Return results
print(empl_debt_corr)
```
The correlation coefficient of `r paste(round(empl_debt_corr, 2))` suggests a very weak positive correlation between the total external debt and the percent of employment within each country, indicating no meaningful connection between these two variables. 

The annual population growth was then compared to the country's debt and another correlation test was performed between these two variables. 
```{r}
# Top 10 countries with highest population growth
top_10_popul <- dbGetQuery(connection, "
    SELECT Country_Name, External_Debt, ROUND(Annual_Population_Growth, 2) AS  Annual_Population_Growth
    FROM international_data
    ORDER BY Annual_Population_Growth DESC
    LIMIT 10
")
# Return results 
print(top_10_popul)

# Debt/Population variable
debt_popul <- dbGetQuery(connection, "
    SELECT External_Debt, Annual_Population_Growth
    FROM international_data;
")

# Remove NA values
debt_popul <- na.omit(debt_popul)

# Correlation test
debt_popul_corr <- cor(debt_popul$External_Debt, debt_popul$Annual_Population_Growth)
# return results
print(debt_popul_corr)

```
The top 10 countries with the highest population growth annually are:

- Niger
- Syrian Arab Republic	
- Singapore	
- Kazakhstan
- Congo, Dem. Rep.	
- Chad
- Angola
- Mali
- Somalia
- Uganda

Given the correlation of `r paste(round(debt_popul_corr, 2))`, there is no correlation between population growth and the debt accumulated by country. 


```{r}
# Debt/Population variable
debt_val <- dbGetQuery(connection, "
    SELECT External_Debt, Debt_Value
    FROM international_data;
")

# Remove NA values
debt_val <- na.omit(debt_val)

# Correlation test
debt_val_corr <- cor(debt_val$External_Debt, debt_val$Debt_Value)
# return results
print(debt_val_corr)
```
```{r}
# scatter plot; external debt by value
ggplot(debt_val, 
       aes(x = External_Debt, y = Debt_Value, color = External_Debt)) +
  # scatter plot
  geom_point() + 
  # Add labels
  labs(title = "External Debt by Value", x = "Total External Debt", y = "External Debt Value")

# disconnect from database
dbDisconnect(connection)

```
Here we can see a strong positive correlation between debt and debt value.
